Dependent Types for Safe and Secure Web Programming

### 2013-11-17 paper/conclusion.tex
Proof reading/trimming pass over the paper

To date,
dependently-typed languages have remained relatively unused for practical
purposes. Advances in embedded domain-specific languages
have allowed many of the advantages of such languages, including the additional
guarantees given by the successful compilation of a program (assuming the use
of specific enough types) to be adopted without additional proofs.

Our framework, through the use of dependent types, provides several static
guarantees. Data submitted by users is inherently unsafe due to the possibility
of attacks such as SQL injection, and this particular threat is ameliorated due
to elements to be associated with specific types during form construction. This
immediately eliminates the possibilities of SQL injection attacks on non-string
types. Since failures are handled transparently, no runtime errors are output
to the browser, meaning that attackers may not use such information to aid
attacks. Additionally, since checking is performed on the types of the form
elements and the types of arguments accepted by the handler, it is impossible
to associate a form with a handler incompatible with the submitted data.

Many external libraries must also adhere to a resource usage protocol, in order to ensure correct operations. Since such libraries do not have as expressive a type system as \idris{}, however, incorrect usage is still possible, for example by developers forgetting to release acquired resources or failing to initialise a library correctly. By creating high-level bindings to these libraries, however, we may statically enforce these resource-usage protocols, ensuring that the libraries are used correctly. Whilst previous work has demonstrated that this is possible through the use of embedded DSLs \cite{brady:edsl} and parameterised algebraic effects \cite{brady:effects}, this paper has provided more substantial examples of real-world applications.

In particular, the framework guarantees that it is not possible for CGI to produce an internal server error due to content being written to the remote host prior to headers. With regard to database access, we may statically guarantee that library calls are made in the correct order, and calls to retrieve rows of data are made only when more data is available. Additionally, by recording any failures in the resource associated with the database effect, we may ensure that any failures at certain stages in the process may be correctly dealt with, and not propagate to further library calls---if a prepared statement fails to compile, for example, the framework guarantees that it will not be executed, without causing a fatal runtime error or memory access violation.

Enforcing resource usage protocols also guards against common programmer errors, saving debugging time by identifying errors at compile time.
\subsection{Further Work}
We have shown that embedded domain-specific languages using dependent types and algebraic effects can be used to increase confidence in web applications by providing additional static guarantees about runtime behaviour, but much more can be done using the same approach.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an interesting method by which external data sources may be used to import external information, such that it may be used during compilation. Through the use of this technique, it becomes possible to make use of the extra type information in order to statically ensure the validity of artefacts such as SQL queries and data structures. If data structures within the program do not conform to a given database schema, for example, then the program will not type-check and as such will produce a compile-time error. 
This is extended to \idris{} through the implementation of dependent type providers \cite{christiansen:dtp}, which makes use of the fact that types can be calculated by functions to avoid generating extra code in the type provider step. Dependent type providers additionally have stronger safety guarantees as they may not generate unchecked code, but at the same time this is matched by a decrease in expressiveness. Nonetheless, such techniques provide a promising mechanism to verify the semantic soundness of programs and we look to investigate their integration in further work.

Dependently-typed languages provide great promise for the construction of secure and correct programs. Through the use of embedded domain-specific languages, we hope that more developers may benefit from the extra guarantees afforded by dependent types, resulting in more stable, secure applications.
% Cryptography bindings would be really good.
% Integration with a web server instead of doing everything over CGI
% Improvements to form handling system
% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors

>>>>

Dependently-typed languages promise to support machine checkable program
correctness proofs, but to date they have remained relatively unused for
practical purposes. By using embedded domain-specific languages, we can
abstract away some of the complexities of creating correctness proofs and
provide expressive libraries, giving guarantees by the successful compilation
of a program (assuming the use of specific enough types) without additional
proofs.

Our framework provides several static
guarantees. Data submitted by users is inherently unsafe and
means systems are vulnerable to attacks
such as SQL injection. This particular threat is ameliorated due
to elements being associated with specific types during form construction. This
immediately eliminates the possibilities of SQL injection attacks on non-string
types. Since failures are handled transparently, no runtime errors are output
to the browser, meaning that attackers may not use such information to aid
attacks. Additionally, since checking is performed on the types of the form
elements and the types of arguments accepted by the handler, it is impossible
to associate a form with a handler incompatible with the submitted data.

Many external libraries also follow (unchecked or dynamically checked) 
resource usage protocols.
Incorrect usage is however still possible, for
example by forgetting to release acquired resources or failing to
initialise a library correctly. By creating high-level bindings to these
libraries, however, we may statically enforce these resource-usage protocols,
ensuring that the libraries are used correctly. Whilst previous work has
demonstrated that this is possible through the use of embedded DSLs
\cite{brady:edsl} and dependent algebraic effects \cite{brady:effects},
this paper has provided more substantial examples of real-world applications.

In particular, the framework guarantees that it is not possible for a CGI 
application to
produce an internal server error due to content being written to the remote
host prior to headers. With regard to database access, we may statically
guarantee that library calls are made in the correct order, and calls to
retrieve rows of data are made only when more data is available. Additionally,
by recording any failures in the resource associated with the database effect,
we may ensure that any failures at certain stages in the process may be
correctly dealt with, and not propagate to further library calls---if a
prepared statement fails to compile, for example, the framework guarantees that
it will not be executed, without causing a fatal runtime error or memory access
violation.
%
Enforcing resource usage protocols also guards against common programmer
errors, saving debugging time by identifying errors at compile time.

\subsection{Further Work}

We have shown that embedded domain-specific languages using dependent types and
algebraic effects can be used to increase confidence in web applications by
providing additional static guarantees about runtime behaviour, but much more
can be done using the same approach.

@@

Type providers, as originally implemented in F\# \cite{msr:tp}, are an
interesting method by which external data sources may be used to import
external information, such that it may be used during compilation. 
In this way, it becomes possible to use the extra type
information to statically ensure the validity of artefacts such as SQL
queries and data structures. If data structures within the program do not
conform to a given database schema, for example, then the program will not
type-check.  This has been implemented for 
\idris{} 
\cite{christiansen:dtp}, exploiting the fact that types can be
calculated by functions to avoid generating extra code in the type provider
step. Dependent type providers additionally have stronger safety guarantees as
they may not generate unchecked code, but at the same time this is matched by a
decrease in expressiveness. Nonetheless, such techniques provide a promising
mechanism to verify the semantic soundness of programs and we look to
investigate their integration in further work.

Dependently-typed languages provide great promise for the construction of
secure and correct programs. Through the use of embedded domain-specific
languages, we hope that more developers may benefit from the extra guarantees
afforded by dependent types, resulting in more stable, secure applications.

% Cryptography bindings would be really good.
% Integration with a web server instead of doing everything over CGI
% Improvements to form handling system
% Less raw SQL, use get more type-safety by using more complex EDSLs for database access thus further minimising errors

### 2013-11-17 paper/effects.tex
Proof reading/trimming pass over the paper

\noindent
That is, \texttt{Get} returns a value of type \texttt{a} without updating
the resource type. \texttt{Put} returns nothing, but has the effect of updating
the resource. In order to make an effect usable, we implement a handler
for a computation context by making an instance of the following class:

@@

\noindent
The \texttt{handle} function takes the input resource, an effect which may
update that resource and execute a side-effect, and a continuation \texttt{k}
which takes the updated resource and the return value of the effect. We use
a continuation here primarily because there is no restriction on the number of
times a handler may invoke the continuation (raising an exception, for example,
will not call the continuation). Reading and updating states in handled
for all computation contexts \texttt{m} as follows:

@@

\end{SaveVerbatim}

@@

A concrete effect is simply an algebraic effect type paired with its current
resource type. This, and other
technical details, are explained in full elsewhere~\cite{brady:effects}.
For the purposes of this paper, it suffices to know how to describe and
handle new algebraic effects.

@@

More generally, a program might modify the set of effects available.
This might be desirable for several reasons, such as adding a new
effect, or to update an index of a dependently typed state. In this
case, we describe programs using the \texttt{EffM} data type:

@@

\noindent
\texttt{EffM} is parameterised over the context and type as before, but
separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
In fact, \texttt{Eff}
is defined in terms of \texttt{EffM}, with equal input/output effects.
We can use this to describe how effects follow resource protocols. A simple
example is a file access protocol, where a file must be opened before it
is read or written, and a file must be closed on exit:

\begin{SaveVerbatim}{fileeff}

data FileIO : Effect where
     Open  : String -> (m : Mode) -> 
             FileIO () (OpenFile m) ()
     Close : FileIO (OpenFile m) () ()

     ReadLine  : FileIO (OpenFile Read)  
                        (OpenFile Read) String
     WriteLine : String -> 
                 FileIO (OpenFile Write) 
                        (OpenFile Write) ()
     EOF       : FileIO (OpenFile Read)  
                        (OpenFile Read) Bool

\end{SaveVerbatim}
\useverb{fileeff}

\noindent
Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file changes an empty resource to
a resource containing an open file; reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions, as well as throwing an exception
if any operation fails.

@@

testFile : Eff IO [FILE_IO (), STDIO] () 
testFile = catch (do open "testFile" Read
                     str <- readLine
                     close
                     putStrLn str)
            (\err => putStrLn ("Error: " ++ show err))

@@

\noindent
The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit.
Therefore, attempting to write to the file, or failing to open or close the
file, would cause a \emph{compile-time} error.

>>>>

\noindent
That is, \texttt{Get} returns a value of type \texttt{a} without updating
the resource type. \texttt{Put} returns nothing, but has the effect of updating
the resource. To make an effect usable, we implement a handler
for a computation context by making an instance of the following class:

@@

\noindent
The \texttt{handle} function takes the input resource, an effect which may
update that resource and execute a side-effect, and a continuation \texttt{k}
which takes the updated resource and the return value of the effect. We use
a continuation here primarily because there is no restriction on the number of
times a handler may invoke the continuation (raising an exception, for example,
will not invoke it). Reading and updating states is handled
for all computation contexts \texttt{m}:

@@

\end{SaveVerbatim}
\useverb{statehandler}

@@

\noindent
A concrete effect is simply an algebraic effect type paired with its current
resource type. This, and other
technical details, are explained in full elsewhere~\cite{brady:effects}.
For the purposes of this paper, it suffices to know how to describe and
handle new algebraic effects.

@@

\begin{SaveVerbatim}{fileeff}
{-                                             { Input resource type }   { Output resource type } { Value } -}

data FileIO : Effect where
     Open      : String -> (m : Mode) -> FileIO     ()                    (Either () (OpenFile m)) ()
     Close     :                         FileIO     (OpenFile m)          ()                       ()

     ReadLine  :                         FileIO     (OpenFile Read)       (OpenFile Read)          String
     WriteLine : String ->               FileIO     (OpenFile Write)      (OpenFile Write)         ()
     EOF       :                         FileIO     (OpenFile Read)       (OpenFile Read)          Bool
\end{SaveVerbatim}

\begin{figure*}[t]
\begin{center}
\useverb{fileeff}
\end{center}
\caption{File Protocol Effect}
\label{fig:fileeffect}
\end{figure*}
More generally, a program might modify the set of effects available.
This might be desirable for several reasons, such as adding a new
effect, or to update an index of a dependently typed state. In this
case, we describe programs using the \texttt{EffM} data type:

@@

\noindent
\texttt{EffM} is parameterised over the context and type as before, but
separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
In fact, \texttt{Eff}
is defined in terms of \texttt{EffM}, with equal input/output effects.
We can use this to describe how effects follow resource protocols. A simple
example is a file access protocol, where a file must be opened before it
is read or written, and a file must be closed on exit.

Figure \ref{fig:fileeffect} shows how the protocol is encoded as an
effect.
Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file may fail, so
changes an empty resource to
a resource containing either a unit or an open file; 
reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions.

@@

testFile : Eff IO [FILE_IO (), STDIO] () 
testFile = do open "testFile" Read
              if_valid then do str <- readLine
                            close
                            putStrLn str
                       else putStrLn "Error"

@@

\noindent
The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit. We use \texttt{if\_valid} to handle possible failure---this is
a function provided by the \texttt{Effects} library which checks whether
a resource of type \texttt{Either a b} indicates failure (\texttt{Left a})
or success (\texttt{Right b}) and branches and updates the resource
accordingly.
Therefore, attempting to write to the file, failing to check for
success, or failing to open or close the
file, would cause a \emph{compile-time} error.

### 2013-11-17 paper/forms.tex
Proof reading/trimming pass over the paper

\label{form}
Programming web applications often involves processing user data, which may
then be used in further effectful computations. Data submitted using a form is
transmitted over the internet as a string as part of an HTTP request, which
traditionally involves losing associated type information.

@@

In this section, we present a mechanism by which we introduce a DSL
for the creation of web forms which preserve type information, implemented
as a dependent algebraic effect. Once the form has
been submitted, retrieved information is passed directly to a
developer-specified function for handling, without the need to manually check
and deserialise data.

We begin with a simple example of a form which requests a user's name, and
echoes it back. Firstly, we define a form handler which echoes back a string
provided by the form handler. It has one argument of type \texttt{Maybe
String}, which accounts for the possibility that the user may have specified
incorrect data within the form.

@@

sayHello : Maybe String -> 
           FormHandler [CGI (InitialisedCGI TaskRunning)]
sayHello (Just name) = output ("Hello, " ++ name ++ "!")
sayHello _ = output "Error!"

@@

We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:

@@

handlers : HandlerList
handlers = [handler args=[FormString], 
                    effects=[CgiEffect], 
                    fn=sayHello, 
                    name="sayHello"]

@@

We also define a form to take in a name from the user, and specify that it should use the \texttt{sayHello} handler.

@@

showHelloForm : UserForm
showHelloForm = do
  addTextBox "Name" FormString Nothing
  useEffects [CgiEffect]
  addSubmit sayHello handlers

@@

In order to specify a form, we once again use \texttt{Effects}. By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: the specification of an
incompatible handler will result in a compile-time type error. The \texttt{Form}
effect and associated resource \texttt{FormRes} is given in Figure 
\ref{fig:formeffect}.
The \texttt{using} notation here indicates that within
the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
the given type.

@@

\noindent
These make use of the resource associated with the effect, \texttt{FormRes}, to
construct the form. Adding a field such as a text box adds a new type, \texttt{fty}
to the list of field types, carried in the resource. When the form is complete,
the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field
types, and using it to construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

\noindent
All forms are required to include a submit button, as mandated by the
requirement that

The input and output resource contains an empty list of types, which means that
any form which includes fields must also include a submit button. Adding fields
adds to the list of types, and only adding a submit button empties that list.
Note that there is no need to restrict this effect to running in the \texttt{IO}
monad since creating a form merely returns HTML text, with no side-effects by
default.

Handlers may only be associated with a form if they have argument types
corresponding to the types associated with the form elements. Additionally, we
wish to name the function in order for it to be serialised, whilst requiring a
proof that the specified name is associated with the function. If this were not
the case, it would be possible to specify a function which satisfies the type
requirement, without guaranteeing that it the serialised data corresponded to
that function, thus rendering the check pointless.

Before associating a handler function with the form, we must specify the
effects available to the handler. This is done through the use of the
\texttt{useEffects}, which updates the list of effects in the type of the form
resource. By doing this, we may subsequently use the effects in calculations at
the type level, in particular when calculating the type of the handler function
for the form.

@@

\begin{SaveVerbatim}{mkhandlerfn}

mkHandlerFn' : List FormTy -> List WebEffect -> Type
mkHandlerFn' [] effs = FormHandler (map webEffect effs) 
mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
                              mkHandlerFn' xs effs

mkHandlerFn : MkHandlerFnTy -> Type 
mkHandlerFn (tys, effs) = mkHandlerFn' tys effs

@@

@@

\noindent
The \texttt{mkHandlerFn} function takes a tuple describing the arguments and
web effects available to the handler function. When constructing the function
type, we encase all arguments withing \texttt{Maybe} type, in order to handle
failure should the supplied data fail to parse to the specified type.
%
To store a reference to a handler function, we use the \texttt{HandlerFn} type:

@@

\noindent
The \texttt{**} notation denotes a dependent pair, where the type of the second
element of the pair is parameterised over the value of the first element. It is
an existential binding;
the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
\texttt{P x} holds''.
Therefore a \texttt{HandlerFn} states that there exists a function type
\texttt{ft} such that we have a handler for it, and a unique string identifier
which is used to serialise a
reference to the handler function.

In order to abstract away from this implementation detail, we make use of
\idris{} syntax rewriting rules. This allows us to define the following syntax
rewrite rule:

@@

\noindent
We may then define handlers in a more intuitive fashion, without being
concerned with the implementation details. This allows us to write a handler
with one String argument, making use of the CGI effect, associated with the
\texttt{sayHello} handler function as follows:

@@

handler args=[FormString], 
        effects=[CgiEffect], 
        fn=sayHello, 
        name="sayHello"

@@

To enforce the requirement that a supplied handler function must reside in the
list of available handlers, and therefore allow us to retrieve the name with
which to serialise the handler, we require a \textit{list membership proof},
\texttt{FnElem f fns}, which statically guarantees that a given item resides in
a list.

@@

\noindent
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the
form elements and the effects used by the handler function. \texttt{FnHere}
is a proof that the element is at the head of the current point of the list,
whereas \texttt{FnThere} is a proof that the element is in the tail of 
the list. 
%TODO: this should probably be rewritten
We then use a simple automated proof search to
automatically generate the proof at compile time, should one exist. The proof
may then be used in subsequent computations: in our case, we use it to retrieve
the unique identifier for the function. If the automated proof search fails,
compilation will fail.

@@

Although sending details of the handler function to the client may appear to be
a security risk, we envisage that the use of symmetric encryption or a
cryptographic nonce would alleviate this. Ultimately, we hope to implement a
web server with persistent state, which would eliminate the need for
serialisation altogether.

Running form construction is achieved as an operation of the CGI
effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
The generated metadata describing the handler function is serialised as a
hidden HTML field.

@@

decEq : DecEq t => (x : t) -> (y : t) -> Dec (x1 = x2)

@@

\noindent
We may then parse the arguments according to the types specified by the handler
function, and then apply the arguments to the handler function through the use
of partial application. Finally, we may run the handler function, ensuring that
all updates made to the CGI state are propagated.

>>>>

\label{form}
Programming web applications often involves processing user data, which may
then be used in further effectful computations. Data submitted using a form is
transmitted as a string as part of an HTTP request, which
traditionally involves losing associated type information.

@@

In this section, we present a DSL
for the creation of web forms which preserve type information, implemented
as a resource-dependent algebraic effect. Once the form has
been submitted, retrieved information is passed directly to a
developer-specified function for handling, without the need to manually check
and deserialise data.

We begin with a simple example of a form which requests a user's name, and
echoes it back. Firstly, we define a form handler which echoes back a string
provided by the form handler. It has one argument of type \texttt{Maybe
String}, which accounts for the possibility that the user may have 
provided invalid data:

@@

echo : Maybe String -> 
     FormHandler [CGI (InitialisedCGI TaskRunning)]
echo (Just name) = output ("Hello, " ++ name ++ "!")
echo _ = output "Error!"

@@

\noindent
We then specify this in a list of handlers, detailing the arguments, available effects, handler function and unique identifier:

@@

handlers : HandlerList
handlers = [handler args=[FormString], 
                    effects=[CgiEffect], 
                    fn=echo, 
                    name="echo"]

@@

\noindent
We also define a form to take in a name from the user, and specify that it
should use the \texttt{echo} handler.

@@

showHelloForm : UserForm
showHelloForm = do
  addTextBox "Name" FormString Nothing
  useEffects [CgiEffect]
  addSubmit echo handlers

@@

\noindent
Again, we use \texttt{Effects} to build a form.
By recording
the type of each form element as it is added in the type of the form, we may
statically ensure that the user-supplied handler function is of the correct
type to handle the data supplied by the form: using an
incompatible handler will result in a compile-time type error. The \texttt{Form}
effect and associated resource \texttt{FormRes} is given in Figure 
\ref{fig:formeffect}.
The \texttt{using} notation here indicates that within
the block, where \texttt{G} and \texttt{E} occur they are implicit arguments with
the given type.

@@

\noindent
These use the resource associated with the effect, \texttt{FormRes}, to
construct the form. Adding a field such as a text box adds a new type,
\texttt{fty} to the list of field types, carried in the resource. When the form
is complete, the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field types, and using it to
construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

\noindent
%All forms are required to include a submit button, as mandated by the
%requirement that 
%
The input and output resource contains an empty list of types, which means that
any form which includes fields must also include a submit button. Adding fields
adds to the list of types, and only adding a submit button empties that list.
Note that there is no need to restrict this effect to running in the \texttt{IO}
monad since creating a form merely returns HTML text, with no side-effects by
default.

Handlers may only be associated with a form if they have argument types
corresponding to the types associated with the form elements. Additionally, we
wish to name the function in order for it to be serialised, whilst requiring a
proof that the specified name is associated with the function. If this were not
required, it would be possible to use a function satisfying the type
requirement, without guaranteeing that the serialised data corresponded to
that function.

Before associating a handler function with the form, we must specify the
effects available to the handler. This is done with
\texttt{useEffects}, which updates the list of effects in the type of the form
resource. By doing this, we may subsequently use the effects in calculations at
the type level, in particular when calculating the type of the handler function
for the form.

@@

\begin{SaveVerbatim}{mkhandlerfnp}

mkHandlerFn' : List FormTy -> List WebEffect -> Type
mkHandlerFn' [] effs = FormHandler (map webEffect effs) 
mkHandlerFn' (x :: xs) effs = Maybe (interpFormTy x) -> 
                              mkHandlerFn' xs effs 
\end{SaveVerbatim}

\begin{SaveVerbatim}{mkhandlerfn}
mkHandlerFn : MkHandlerFnTy -> Type 
mkHandlerFn (tys, effs) = mkHandlerFn' tys effs

@@

\useverb{mkhandlerfnp}

@@

\noindent
The \texttt{mkHandlerFn} function takes a tuple describing the arguments and
web effects available to the handler function. When constructing the function
type, we wrap all arguments in a \texttt{Maybe}, in order to handle
failure should the supplied data fail to parse as the required type.
%
To store a reference to a handler function, we use the \texttt{HandlerFn} type:

@@

\noindent
The \texttt{**} notation denotes a dependent pair, in which the type of the second
element of the pair is parameterised over the value of the first element. It is
an existential binding;
the notation \texttt{(x ** P x)} can be read as ``there exists an \texttt{x} such that 
\texttt{P x} holds''.
Therefore a \texttt{HandlerFn} states that there exists a function type
\texttt{ft} such that we have a handler for it, and a unique string identifier
which is used to serialise a
reference to the handler function.

In order to abstract away from this implementation detail, we make use of
\idris{} syntax rewriting rules. This allows us to define the following:

@@

\noindent
We may then define handlers in a more intuitive fashion, without being
concerned with the implementation details. This allows us to write a handler
with one String argument, making use of the CGI effect, associated with the
\texttt{echo} handler function as follows:

@@

handler args=[FormString], 
        effects=[CgiEffect], 
        fn=echo, 
        name="echo"

@@

To enforce the requirement that a supplied handler function must be in the
list of available handlers, and therefore allow us to retrieve the name with
which to serialise the handler, we require a \textit{list membership proof},
\texttt{FnElem f fns}, which statically guarantees that a given item resides in
a list.

@@

\noindent
\texttt{FnElem} is parameterised over \texttt{G} and \text{E}, the types of the
form elements and the effects used by the handler function. \texttt{FnHere}
is a proof that the element is at the head of the current point of the list,
whereas \texttt{FnThere} is a proof that the element is in the tail of 
the list. 
%TODO: this should probably be rewritten
We then use an automatic proof search to
generate the proof at compile time, should one exist. The proof
may then be used in subsequent computations: we use it to retrieve
the unique identifier for the function. If the automated proof search fails,
compilation will fail.

@@

Although sending details of the handler function to the client may appear to be
a security risk, we envisage that the use of symmetric encryption or a
cryptographic nonce would mitigate this. Ultimately, we hope to implement a
web server with persistent state, which would eliminate the need for
serialisation altogether.

Running form construction is achieved as an operation of the \texttt{CGI}
effect, \texttt{AddForm}, which then outputs the generated HTML to the page.
The generated metadata describing the handler function is serialised as a
hidden HTML field.

@@

decEq : DecEq t => (x : t) -> (y : t) -> Dec (x = y)

@@

\noindent
We may then parse the arguments according to the types specified by the handler
function, and then apply the arguments to the handler function.
Finally, we may run the handler function, ensuring that
all updates made to the CGI state are propagated.

### 2013-11-17 paper/introduction.tex
Proof reading/trimming pass over the paper

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. A significant drawback, however, is that such languages
do not provide  the same static guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

@@

\noindent
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it. However, this would cause
problems with later steps.

@@

\noindent
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
The difficulty is 
that automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

This complexity can be ameliorated through the use of \textit{embedded
domain-specific languages} (EDSLs), which aim to abstract away the
complexity of the underlying type system. EDSLs allow domain experts to
write verified domain-specific code, with the EDSL itself providing the implicit proof that the written code is correct.

@@

\begin{itemize}
\item A form-handling mechanism, which preserves type information and
manages user input, therefore
increasing applications' resilience to attacks such as SQL injection and cross-site scripting.

\item Representations of CGI, Databases and sessions as
\textit{resource-dependent algebraic effects}, allowing programs to be accepted only when they
follow clearly defined resource usage protocols.

\item A message board application, demonstrating the usage of the framework.

@@

We structure the remainder of this paper as follows. We provide a brief
overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
how this may be used to ensure adherence to resource usage protocols for CGI,
SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
type-safe form handling in Section ~\ref{form}, implemented using
\texttt{Effects}; and discuss the larger example
of a message board system making use of these components in Section
~\ref{messageboard}.

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/idris-hackers/IdrisWeb}.
% =================================================
% =================================================

>>>>

Many web applications are written in dynamically-checked scripting languages
such as PHP, Ruby or Python, which facilitate rapid development
\cite{w3techs:webpls}. However, such languages
do not provide the same static guarantees about runtime behaviour afforded by
programs with more expressive, static type systems, instead relying on
extensive unit testing to ensure correctness and security.

@@

\noindent
We should then check to see if the connection was successful, and exit
if not:
%This check is optional, so
%it would be possible to omit it. However, this would cause
%problems with later steps.

@@

\noindent
Even in this small example, there exists a precise resource usage protocol
which must be followed for successful and robust operation.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

Problems may arise if the protocol is not followed correctly.
A developer may, for example, accidentally close a statement whilst still
retrieving rows, which would cause a runtime error. Similarly, a programmer may
omit closing the statement or connection, which can lead to
problems such as resource leaks in longer-running server applications.
However, in conventional programming languages, there is no way to check
automatically, at compile-time, that a protocol is followed.

In contrast, the use of \textit{dependent types} makes it possible
to specify a program's behaviour precisely, and to check that a 
specification is followed.
%
Unfortunately, automatic verification by a compiler can be difficult or
often impossible, requiring additional proofs to be given by the programmer.

This complexity can be addressed through the use of \textit{embedded
domain-specific languages} (EDSLs) to abstract away the complexity of the
underlying type system. Embedding a domain-specific language in
a dependently typed host language
allows domain experts to write 
domain-specific code, with the EDSL itself used to provide the proof that the
code is correct.

@@

\begin{itemize}
\item Representations of CGI, Databases and sessions as
\textit{resource-dependent algebraic effects}, allowing programs to be accepted
only when they follow clearly defined resource usage protocols. 
(Section~\ref{rup})

\item Type-safe form handling, preserving type information and managing
user input, therefore increasing applications' resilience to attacks such as
SQL injection and cross-site scripting. (Section~\ref{form})

\item An extended example: a message board application, demonstrating the usage
of the framework in practice. (Section~\ref{effects})

@@

We achieve these without extending the host language. Every
resource protocol we implement is pure \Idris{}, using a library for
resource-dependent algebraic effects~\cite{brady:effects} and \Idris{}'
features for supporting domain-specific language implementation such as
syntax macros and overloading. In particular, this means the same techniques
can be applied to other resources, and most importantly, combinations
of resources and DSLs implemented in this way are \emph{composable}.

%We structure the remainder of this paper as follows. We provide a brief
%overview of the \texttt{Effects} framework in Section ~\ref{effects}; explain
%how this may be used to ensure adherence to resource usage protocols for CGI,
%SQLite and a session handler in Section ~\ref{rup}; describe an EDSL for
%type-safe form handling in Section ~\ref{form}, implemented using
%\texttt{Effects}; and discuss the larger example
%of a message board system making use of these components in Section
%~\ref{messageboard}.

The code used to implement the framework and all associated examples used in
this paper is available online at
\url{http://www.github.com/idris-hackers/IdrisWeb}.

% =================================================
% =================================================

### 2013-11-17 paper/main.tex
Proof reading/trimming pass over the paper

\begin{abstract}
Dependently-typed languages allow precise types to be used during development,
facilitating easier reasoning about the operation of programs. However,
stronger types bring a disadvantage that it becomes increasingly difficult to
write programs that are accepted by a type checker and additional proofs may
have to be specified by a programmer.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in a high-level languages which uses
dependent types to enforce certain invariants without additional proof
obligations.

>>>>

\begin{abstract}
Dependently-typed languages allow precise types to be used during development,
facilitating reasoning about programs. However,
stronger types bring a disadvantage that it becomes increasingly difficult to
write programs that are accepted by a type checker and additional proofs may
have to be specified by a programmer.

Embedded domain-specific languages (EDSLs) can help address this problem by
introducing a layer of abstraction over more precise underlying types,
allowing domain-specific code to be written in a high-level language which uses
dependent types to enforce invariants without imposing additional proof
obligations on an application programmer.
